Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
add general Bayesian refs

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitoni2011], lack. We test the robustness of this method using down-sampled, simulated temperature gradients.

>>>>

Here, we present a novel Bayesian hierarchical model [see e.g. @Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitoni2011], lack. We test the robustness of this method using down-sampled, simulated temperature gradients.
